Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-03-31 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
hs: work on eft section, new refs

\begin{table}[th]
\begin{tabular}{c | c c}
\hline \hline
Theory & Constraint & This work \\
\hline
EdGB        & $\ell_{\rm GB} \leqslant 1.18$~km~(GW)~\cite{Lyu:2022gdr} & -- \\
dCS         & $\ell_{\rm CS} \leqslant 8.5$~km~(EM+GW)~\cite{Silva:2020acr}  & $\ell_{\rm CS} \leqslant 38.7$~km \\
cubic EFT   & -- & $\ell_{\rm cEFT} \leqslant 38.2$~km \\
quartic EFT & $\ell_{\rm qEFT} = \dots$~km~\cite{Sennett:2019bpc}  & $\ell_{\rm qEFT} \leqslant 51.3$~km \\
\hline \hline
\end{tabular}
\caption{\hs{Mention it in the Introduction later, in a sort of `executive summary'.}}
\label{tab:bound_summary}
\end{table}

@@

BHs in EdGB gravity have scalar hair, to which we can associate a (monopole) scalar charge $Q$,
related to the asymptotic $1/r$ fall-off of the scalar field.
%
This charge is not an independent parameter, and depends on the BH's mass and spin, %~\cite{Berti:2018cxi},
thus being a ``secondary hair''~\cite{Coleman:1991ku,Herdeiro:2015waa}.
%
Since the scalar field is coupled to a curvature scalar, the $Q$ is larger (smaller)
the smaller (larger) the BH's mass is.

@@

This theory is described by the action~\cite{Jackiw:2003pm},
%
\begin{equation} \label{eq:action_dcs}
    S_{\rm CS} = \frac{1}{16 \pi}
    \int \dV
    \left[
    R - \tfrac{1}{2}(\nabla \vartheta)^2
    + \tfrac{1}{4} \ell^{2}_{\rm CS} \, \vartheta \, {}^{*}RR
    \right],
\end{equation}
%
where $\vartheta$ is a pseudoscalar field, which couples to the Pontryagin
density
${}^{*}RR = {}^{*}R^{\mu}{}_{\nu}{}^{\rho\sigma} R^{\nu}_{\mu\rho\sigma}$,
%
% \begin{equation}
%     {}^{*}RR = {}^{*}R^{\mu}{}_{\nu}{}^{\rho\sigma} R^{\nu}_{\mu\rho\sigma},
% \end{equation}
%
where ${}^{*}R^{\mu}{}_{\nu}{}^{\rho\sigma}$ is the dual of the Riemann tensor,
defined as
%
${}^{*}R^{\mu}{}_{\nu}{}^{\rho\sigma} =
\epsilon^{\mu}{}_{\nu\gamma\delta}
R^{\gamma\delta\rho\sigma} / 2$,
%
and where $\epsilon^{\mu}{}_{\nu\gamma\delta}$ is the Levi-Civita tensor.
%
As in scalar-Gauss-Bonnet theories, $\int \dV {}^{*}RR$ alone is a boundary term~\cite{Jackiw:2003pm},
hence cannot contribute to the field equations in four-dimensions.
%
However, this terms can modify the theory's field equation by coupling it to
$\vartheta$.
%
The coupling strength set by the coupling constant $\ell_{\rm CS}$, with
dimensions of length.

The theory admits same the garden-variety Schwarzschild BH solution of
GR~\cite{Jackiw:2003pm}, but admits non-Kerr solutions for spinning BHs~\cite{Yunes:2009hc,Konno:2009kg}.
%
These BHs carry secondary scalar hair, which, unlike in EdGB gravity, is
dipolar.
%
As a consequence, the leading-scalar radiative channel of BBH to lose energy is quadrupolar,
which affects the GW phase at 2PN order.
%
Deviations from GR at this PN order are poorly constrained with present GW
observations~\cite{LIGOScientific:2021sio}, making dCS gravity unconstrained by
analyses that rely on the inspiral portion of GWs from BBH
binaries~\cite{Nair:2019iur,Perkins:2021mhb}.
%
However, the theory has been constrained by Ref.~\cite{Silva:2020acr}, that found
%
$\ell_{\rm CS} \leqslant 8.5$~km,
%
by folding data from the x-ray observations of the isolated neutron star (NS)
PSR~J0030+0451~\cite{Lommen:2000yt,NANOGrav:2017wvv} by
NICER~\cite{Riley:2019yda,Miller:2019cac} and from the GW observation of the binary
neutron star (BNS) GW170817~\cite{TheLIGOScientific:2017qsa}.

@@

Our last example of modified gravity theories are the so-called effective field theories (EFT) of GR.
They are described by the action
%
\begin{equation} \label{eq:action_eft}
    S_{\rm EFT} = \frac{1}{16 \pi}
    \int \dd^4x \sqrt{-g}
    \left[ R
    +
    \sum_{n \geqslant 2} \ell^{2n - 2} L^{(2n)}
    \right]\,,
\end{equation}
%
where $\ell$ is a lengthscale, assumed to be small compares to the lengthscale
associated with a BH, i.e., $\ell / M \ll 1$, and $L^{(2n)}$ are corrections to
that introduce higher-order curvature tensors (with $2n$ metric derivatives).

>>>>

\begin{table}[th]
\begin{tabular}{c | c c}
\hline \hline
Theory & Constraint & This work \\
\hline
EdGB        & $\ell_{\rm GB} \leqslant 1.18$~km~(GW)~\cite{Lyu:2022gdr} & -- \\
dCS         & $\ell_{\rm CS} \leqslant 8.5$~km~(EM+GW)~\cite{Silva:2020acr}  & $\ell_{\rm CS} \leqslant 38.7$~km \\
cubic EFT   & -- & $\ell_{\rm cEFT} \leqslant 38.2$~km \\
quartic EFT & $\ell_{\rm qEFT} \leqslant 150$~km~\cite{Sennett:2019bpc}  & $\ell_{\rm qEFT} \leqslant 51.3$~km \\
\hline \hline
\end{tabular}
\caption{\hs{Mention it in the Introduction later, in a sort of `executive summary'.}}
\label{tab:bound_summary}
\end{table}

@@

BHs in EdGB gravity have scalar hair, to which we can associate a (monopole) scalar charge $Q$,
related to the asymptotic $r^{-1}$ fall-off of the scalar field.
%
This charge is not an independent parameter, and depends on the BH's mass and spin, %~\cite{Berti:2018cxi},
thus being a ``secondary hair''~\cite{Coleman:1991ku,Herdeiro:2015waa}.
%
Since the scalar field is coupled to a curvature scalar, the $Q$ is larger (smaller)
the smaller (larger) the BH's mass is.

@@

This theory is described by the action~\cite{Jackiw:2003pm,Alexander:2009tp},
%
\begin{equation} \label{eq:action_dcs}
    S_{\rm CS} = \frac{1}{16 \pi}
    \int \dV
    \left[
    R - \tfrac{1}{2}(\nabla \vartheta)^2
    + \tfrac{1}{4} \ell^{2}_{\rm CS} \, \vartheta \, {}^{*}RR
    \right],
\end{equation}
%
where $\vartheta$ is a pseudoscalar field, which couples to the Pontryagin
density
${}^{*}RR = {}^{*}R^{\mu}{}_{\nu}{}^{\rho\sigma} R^{\nu}_{\mu\rho\sigma}$,
%
% \begin{equation}
%     {}^{*}RR = {}^{*}R^{\mu}{}_{\nu}{}^{\rho\sigma} R^{\nu}_{\mu\rho\sigma},
% \end{equation}
%
where ${}^{*}R^{\mu}{}_{\nu}{}^{\rho\sigma}$ is the dual of the Riemann tensor,
defined as
%
${}^{*}R^{\mu}{}_{\nu}{}^{\rho\sigma} =
\epsilon^{\mu}{}_{\nu\gamma\delta}
R^{\gamma\delta\rho\sigma} / 2$,
%
and where $\epsilon^{\mu}{}_{\nu\gamma\delta}$ is the Levi-Civita tensor.
%
As in scalar-Gauss-Bonnet theories, $\int \dV {}^{*}RR$ alone is a boundary term~\cite{Jackiw:2003pm},
hence cannot contribute to the field equations in four-dimensions.
%
However, this terms can modify the theory's field equation by coupling it to
$\vartheta$.
%
The coupling strength set by the coupling constant $\ell_{\rm CS}$, with
dimensions of length.

The theory admits same the garden-variety Schwarzschild BH solution of
GR~\cite{Jackiw:2003pm}, but admits non-Kerr solutions for spinning BHs~\cite{Yunes:2009hc,Konno:2009kg}.
%
These BHs support a scalar field which asymptotically falls-off as $r^{-2}$, to which we can associate
a scalar dipole charge.
%
As a consequence, the leading-scalar radiative channel of BBH to lose energy is quadrupolar,
which affects the GW phase at 2PN order.
%
Deviations from GR at this PN order are poorly constrained with present GW
observations~\cite{LIGOScientific:2021sio}, making dCS gravity unconstrained by
analyses that rely on the inspiral portion of GWs from BBH
binaries~\cite{Nair:2019iur,Perkins:2021mhb}.
%
However, the theory has been constrained by Ref.~\cite{Silva:2020acr}, that found
%
$\ell_{\rm CS} \leqslant 8.5$~km,
%
by folding data from the x-ray observations of the isolated neutron star (NS)
PSR~J0030+0451~\cite{Lommen:2000yt,NANOGrav:2017wvv} by
NICER~\cite{Riley:2019yda,Miller:2019cac} and from the GW observation of the binary
neutron star (BNS) GW170817~\cite{TheLIGOScientific:2017qsa}.

@@

Our last example of modified gravity theories are the so-called effective field
theories (EFT) of GR~\cite{Endlich:2017tqa,Sennett:2019bpc,deRham:2020ejn,Cano:2020cao,Cano:2021myl}.
%
They are described by the action
%
\begin{equation} \label{eq:action_eft}
    S_{\rm EFT} = \frac{1}{16 \pi}
    \int \dd^4x \sqrt{-g}
    \left[ R
    +
    \sum_{n \geqslant 2} \ell^{2n - 2} L^{(2n)}
    \right]\,,
\end{equation}
%
where $\ell$ is a lengthscale, assumed to be small compares to the lengthscale
associated with a BH, i.e., $\ell / M \ll 1$, and $L^{(2n)}$ are corrections to
that introduce higher-order curvature tensors (with $2n$ metric derivatives).
